### 2014-08-25 doc/thesis/typedlua.tex
Fixing the syntax of table types in the section about tables and interfaces

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table:
\begin{align}
\label{syn:table1}
\texttt{\{k:t\}} & \equiv \texttt{\{k:t|nil\}}\\
\label{syn:table2}
\texttt{\{t\}} & \equiv \texttt{\{number:t|nil\}}\\
\label{syn:table3}
\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}} & \equiv
\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}
\end{align}
The first syntax defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type \texttt{\{k:t|nil\}}.
This table type represents a map that maps values of type
\texttt{k} to values of type \texttt{t|nil},
as Lua returns \texttt{nil} when we use a non-existing key to index a table.
The second syntax defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type \texttt{\{number:t|nil\}}.
This table type is just a syntactic sugar to the definition of a
map that maps values of type \texttt{number} to values of type \texttt{t|nil}.
The third syntax defines table types for records: it is written
$\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
and maps to the same table type.
This table type represents a record that maps the literal types
$\texttt{k}_{\texttt{i}}$\texttt{, ..., }$\texttt{k}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.

@@

We can also use the style of the inferred type to define heterogeneous
tuples:
\begin{verbatim}
    local album = { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

@@

If our record type has optional fields, that is, fields that can be \texttt{nil},
we need to use an explicit type declaration when declaring a
variable of this record type, as we show in the following example:
\begin{verbatim}
    local interface Person
      firstname:string
      middlename:string?
      lastname:string
    end
    local user1:Person = { firstname = "Lewis",
                           middlename = "Allan" or nil,
                           lastname = "Reed" }
    local user2:Person = { lastname = "Reed",
                           firstname = "Lou" }
\end{verbatim}

We need an explicit type declaration because neither of the types
that the Typed Lua compiler assigns to the above table constructors
is a subtype of
\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
the type that \texttt{Person} describes.
While declaring \texttt{user1}, we use \texttt{or} to cast the type of
\texttt{middlename} because of the contravariance in the type of
mutable fields.

@@

The following example shows the description file for \texttt{lmd5},
a MD5 digest library for Lua that is written in C:
\begin{verbatim}
    userdata md5_context
      __tostring : (self) -> (string)
      clone : (self) -> (self)
      digest : (self|string, value?) -> (string)
      new : () -> (self)
      reset : (self) -> (self)
      update : (self, string*) -> (self)
      version : string
    end

>>>>

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table:
\begin{align}
\label{syn:table1}
\texttt{`\{'} \; \textit{keytype} \; \texttt{`:'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table2}
\texttt{`\{'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table3}
\texttt{`\{'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type} \;
\{\texttt{`,'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}\} \;
[\texttt{`,'} \; \textit{type}] \; \texttt{`\}'}
\end{align}
The syntax \ref{syn:table1} defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type \texttt{\{k:t|nil\}}.
This table type represents a map that maps values of type
\texttt{k} to values of type \texttt{t|nil},
as Lua returns \texttt{nil} when we use a non-existing key to index a table.
The syntax \ref{syn:table2} defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type \texttt{\{number:t|nil\}}.
This table type is just a syntactic sugar to the definition of a
map that maps values of type \texttt{number} to values of type \texttt{t|nil}.
The syntax \ref{syn:table3} defines table types for records with an
optional array part: it is written
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
and maps to the same table type.
This table type represents a record that maps the literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.

@@

We can also use the style of the inferred type to define heterogeneous
tuples:
\begin{verbatim}
    local album:{1:string, 2:number, 3:string} =
      { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

@@

If our record type has optional fields, that is, fields that can be \texttt{nil},
we need to use an explicit type declaration when declaring a
variable of this record type, as we show in the following example:
\begin{verbatim}
    local interface Person
      firstname:string
      middlename:string?
      lastname:string
    end
    local user1:Person = { firstname = "Lewis",
                           middlename = "Allan",
                           lastname = "Reed" }
    local user2:Person = { lastname = "Reed",
                           firstname = "Lou" }
\end{verbatim}

We need an explicit type declaration because neither of the types
that the Typed Lua compiler assigns to the above table constructors
is a subtype of
\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
the type that \texttt{Person} describes.

@@

The following example shows the description file for \texttt{lmd5} \citep{lmd5},
a MD5 digest library for Lua that is written in C:
\begin{verbatim}
    userdata md5_context
      __tostring : (self) -> (string)
      clone : (self) -> (self)
      digest : (self|string, value?) -> (string)
      new : () -> (self)
      reset : (self) -> (self)
      update : (self, string*) -> (self)
      version : string
    end
